---
title: "Getting started with distanamo"
author: "Matthieu Viry"
date: '`r Sys.Date()`'
output: 
  rmarkdown::html_vignette:
    fig_width: 5
    toc: true
    toc_depth: 1
    number_sections: true
vignette: >  
  %\VignetteIndexEntry{Getting started with distanamo}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

## Introduction

This package allows you to create distance cartograms (or distance anamorphoses, hence the name).

Distance cartograms are a type of cartogram that deforms the layers of a map according to the distances
between a set of source points and a set of image points.

This is done by extending (by interpolation) to the layer(s) of the study area (territorial divisions, network...) the
local displacement between the source coordinates and the image coordinates, derived from the distances between each pair
of homologous points (source / image points).

The relation between the source points and the image points, and thus the relative position of the image points compared
to the source points, must depend on the studied theme (such as positions in access time for which this package provides
some helper functions to generate the image points from the durations between the points).


## Usage

### Basics
To use this package you need to provide two sets of homologous points : *source points* and *image points*.
They are used to create an interpolation grid that will be used to deform the layer(s) of interest.

```{r, fig.width=5,fig.height=5, fig.show='hold'}
library(sf)
library(distanamo)
# Read source points, image points and the background layer to deform
source_pts <- st_read(
  dsn = system.file("gpkg/data-prefecture.gpkg", package = "distanamo"),
  layer = 'prefecture', quiet = TRUE
)
image_pts <- st_read(
  dsn = system.file("gpkg/data-prefecture.gpkg", package = "distanamo"),
  layer = 'image-points', quiet = TRUE
)
background_layer <- st_read(
  dsn = system.file("gpkg/data-prefecture.gpkg", package = "distanamo"),
  layer = 'departement', quiet = TRUE
)

# Create the interpolation grid
igrid <- dc_create(
  source_points = source_pts, 
  image_points = image_pts, 
  precision = 2, 
  bbox = st_bbox(background_layer)
)

# Use it to deform our layer of interest
deformed_background <- dc_interpolate(igrid, background_layer)

# Display the deformed layer
plot(st_geometry(deformed_background))
```

```{r, fig.width=10,fig.height=10, fig.show='hold'}
# Display useful information
summary(igrid)

# Plot information about the interpolation grid
par(mfrow = c(2, 2))
plot(igrid, ask = FALSE)
par(mfrow = c(1, 1))
```

### Adjusting the image points to the source points

In some cases, you may want to adjust the image points to the source points using an affine or a Euclidean transformation.

For example, here, if the image points represent locations in spatial cognition (and thus are not directly comparable to
the source points, aren't in the same coordinate system, etc.), you need to adjust them to the source points.

```{r}
pos_result <- dc_adjust(
    source_points = source_pts,
    image_points = image_pts,
    adjustment_type = "euclidean"
)

# Create the interpolation grid
igrid <- dc_create(
  source_points = pos_result$source_points,
  image_points = pos_result$image_points,
  precision = 2.0, 
  bbox = st_bbox(background_layer)
)
```


This is also what is done internally when using the `dc_generate_positions_from_durations` function (see below)
after it has performed a Principal Coordinates Analysis (PCoA) on the duration matrix.

### Generating image points from a reference point and durations from the reference point to all the other points

Optionally you can provide a layer of source points and matrix of duration between
the points.

This duration matrix will be used to extract the duration between a reference point
and all the other points, allowing to use the `dc_move_from_reference_point` function to move closer / farther
points from the reference point depending on if they can be reached faster or slower of
the mean speed (between the reference point and all the others).

The cartogram obtained by this method qualifies the *unipolar accessibility* of a location
(the reference point used in the `dc_move_from_reference_point` function).
It's also sometimes referred to as a *centered time cartogram*.

```{r, fig.width=5,fig.height=5, fig.show='hold'}
# Read source points and layer to be deformed
source_pts <- st_read(
  dsn = system.file("gpkg/data-prefecture.gpkg", package = "distanamo"),
  layer = 'prefecture', quiet = TRUE
)
background_layer <- st_read(
  dsn = system.file("gpkg/data-prefecture.gpkg", package = "distanamo"),
  layer = 'departement', quiet = TRUE
)
bbox <- sf::st_bbox(background_layer)

# Read durations between points
d <- read.csv(system.file("csv/mat.csv", package = "distanamo"), row.names = 1)
d[1:5, 1:5]

dv <- d['GRENOBLE', ]
# So we have only the duration between GRENOBLE and all the other points
dv[1:5]

source_pts$durations <- as.double(dv)

ref_point <- subset(source_pts, source_pts$NOM_COM == "GRENOBLE")
other_points <- subset(source_pts, !source_pts$NOM_COM == "GRENOBLE")

# Move points to create the image points layer
positioning_result <- dc_move_from_reference_point(
  reference_point = ref_point,
  other_points = other_points,
  duration_col_name = "durations",
  factor = 1
)

# Create the interpolation grid
igrid <- dc_create(
  positioning_result$source_points,
  positioning_result$image_points,
  2.0,
  bbox
)

# Deform the target layer
deformed_background <- dc_interpolate(igrid, background_layer)

# Display the deformed layer
plot(st_geometry(deformed_background))
```


A popular way of representing this type of cartogram is to add concentric circles (separated by a constant time) around
the reference point.
This can be done using the `dc_concentric_circles` function and the result of the `dc_move_from_reference_point` function
(note that the steps parameter is the time, in the same unit as the durations, between each circle).


```{r, fig.width=5,fig.height=5, fig.show='hold'}
circles <- dc_concentric_circles(
  positioning_result,
  steps = list(60, 120, 180)
)
plot(st_geometry(deformed_background))
plot(st_geometry(circles), add = TRUE)
```


### Generating image points from a durations matrix between all the points

Optionally you can provide a matrix of durations between all the points as well as the positions of the source points
and use the `dc_generate_positions_from_durations` function to generate the image points.

This function will perform Principal Coordinates Analysis (PCoA, a form a Multidimensional scaling) on the duration matrix
to generate the positions of the points in a 2D space.
It will then adjust these points (using an affine or a Euclidean transformation) to the source points to generate the
final image points that can be used to create the interpolation grid.

The cartogram obtained by this method qualifies the *global accessibility* (or the *multipolar accessibility*)
of a territory.
```{r, fig.width=5,fig.height=5, fig.show='hold'}
# Read source points and layer to be deformed
source_pts <- st_read(
  dsn = system.file("gpkg/data-prefecture.gpkg", package = "distanamo"),
  layer = 'prefecture', quiet = TRUE
)
background_layer <- st_read(
  dsn = system.file("gpkg/data-prefecture.gpkg", package = "distanamo"),
  layer = 'departement', quiet = TRUE
)

# Read durations between points
d <- read.csv(system.file("csv/mat.csv", package = "distanamo"), row.names = 1)

pos_result <- dc_generate_positions_from_durations(d, source_pts)

# Display useful information about the result of the positioning
summary(pos_result)
plot(pos_result)
```

```{r, fig.width=10,fig.height=10, fig.show='hold'}
# Create the interpolation grid
igrid <- dc_create(
  pos_result$source_points,
  pos_result$image_points,
  2.0,
  sf::st_bbox(background_layer)
)

summary(igrid)
par(mfrow=c(2,2))
plot(igrid, ask = FALSE)
par(mfrow=c(1,1))
```


```{r, fig.width=5,fig.height=5, fig.show='hold'}
# Deform the target layer
deformed_background <- dc_interpolate(igrid, background_layer)

plot(st_geometry(deformed_background))
```


### Deforming multiple layers at once

If you want to deform multiple layers in parallel with an interpolation grid, you can use the `dc_interpolate_parallel`
function.

```R
result_layers <- dc_interpolate_parallel(
  igrid,
  list(layer1, layer2, layer3)
)
```


## More information about the method and references 

This package is a port of the **[Darcy](https://thema.univ-fcomte.fr/productions/software/darcy/)** standalone software regarding the bidimensional regression and the backgrounds layers deformation.  
All credit for the contribution of the method goes to **Colette Cauvin** *(Théma - Univ. Franche-Comté)* and for the
reference Java implementation goes to **Gilles Vuidel** *(Théma - Univ. Franche-Comté)*.

The method is also available as a **QGIS plugin** ([GitHub repository](https://github.com/mthh/QgisDistanceCartogramPlugin) / [QGIS plugin repository](https://plugins.qgis.org/plugins/dist_cartogram/)).

This R package is a wrapper around the Rust library [`distance-cartogram-rs`](https://github.com/mthh/distance-cartogram-rs) which can be used directly from Rust.

### About the method

- Cauvin, C. (2005). A systemic approach to transport accessibility. A methodology developed in Strasbourg: 1982-2002. Cybergeo: European Journal of Geography. DOI: [10.4000/cybergeo.3425](https://doi.org/10.4000/cybergeo.3425).

- Cauvin, C., & Vuidel, G. (2009). Darcy 2.0 - Mode d'emploi (https://thema.univ-fcomte.fr/productions/software/darcy/download/me_darcy.pdf).

- Tobler, W. R. (1994). Bidimensional regression. Geographical Analysis, 26(3), 187-212. DOI: [10.1111/j.1538-4632.1994.tb00320.x](https://doi.org/10.1111/j.1538-4632.1994.tb00320.x)

- Bronner, A. C. (2023).  Cartogrammes, anamorphoses : des territoires transformés. In Traitements et cartographie de l’information géographique, ISTE Group (pp.231-271). DOI: [10.51926/ISTE.9161.ch7](https://doi.org/10.51926/ISTE.9161.ch7).


### About distance (or time) cartograms in general: their usability, the other methods to create them, etc.

- Ullah, R., Mengistu, E., van Elzakker, C. & Kraak, M. (2016). Usability evaluation of centered time cartograms. Open Geosciences, 8(1), 337-359. DOI: [10.1515/geo-2016-0035](https://doi.org/10.1515/geo-2016-0035).

- Hong, S., Kim, Y. S., Yoon, J. C., & Aragon, C. (2014). Traffigram: distortion for clarification via isochronal cartography. In Proceedings of the SIGCHI Conference on Human Factors in Computing Systems (pp. 907–916). Association for Computing Machinery. DOI: [10.1145/2556288.2557224](https://doi.org/10.1145/2556288.2557224).

- Ullah, R., Kraak, M. J., & Van Elzakker, C. (2013). Using cartograms to explore temporal data: do they work. GeoViz, 2013.

- Ullah, R., & Kraak, M. J. (2014). An alternative method to constructing time cartograms for the visual representation of scheduled movement data. Journal of Maps, 11(4), 674–687. DOI: [10.1080/17445647.2014.935502](https://doi.org/10.1080/17445647.2014.935502).

- Hong, S., Kocielnik, R., Min-Joon Yoo, Battersby, S., Juho Kim, & Aragon, C. (2017). Designing interactive distance cartograms to support urban travelers. In 2017 IEEE Pacific Visualization Symposium (PacificVis) (pp. 81-90).

- Shimizu, E. and Inoue, R. (2009). A new algorithm for distance cartogram construction. International Journal of Geographical Information Science 23(11): 1453-1470. DOI: [10.1080/13658810802186882](https://doi.org/10.1080/13658810802186882).

