---
title: "Creating a distance cartogram using OSM data"
author: "Matthieu Viry"
date: '`r Sys.Date()`'
output: 
  rmarkdown::html_vignette:
    fig_width: 14
    toc: true
    toc_depth: 1
    number_sections: true
vignette: >  
  %\VignetteIndexEntry{Creating a distance cartogram using OSM data}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 7
)
```


## Introduction

We are looking to build a unipolar distance cartogram (also called a *centered time cartogram*) using only data from OpenStreetMap (OSM): it is going to use both geographic layers describing the territory and travel time data from OSM.

This type of cartogram visualizes travelling times from a reference starting location to other destinations in a selected area. In the resulting map, the geographic distance are replaced by time distance and the underlying map layers are deformed accordingly.

This highlights the use of several packages developed at [RIATE](https://github.com/riatelab), including :

- `mapsf` to create maps with `sf` objects,
- `osrm` for calculating distance-time matrices from OSM data,
- `maposm` to easily retrieve OSM data (this is a wrapper around `osmdata`),
- `distanamo` for creating distance cartograms.

## Installing and importing packages

`mapsf` and `osrm` are available both on CRAN and on [riatelab's r-universe](https://riatelab.r-universe.dev/builds) while `distanamo` and `maposm` are only available from [riatelab's r-universe](https://riatelab.r-universe.dev/builds).

As a result, we propose to install all the packages from r-universe:

```{r}
repos <- c('https://riatelab.r-universe.dev', 'https://cloud.r-project.org')

install.packages("mapsf", repos = repos)
install.packages("osrm", repos = repos)
install.packages('distanamo', repos = repos)
install.packages('maposm', repos = repos)
```

Once installed, we can import the necessary packages:

```{r}
library(sf)
library(mapsf)
library(osrm)
library(maposm)
library(distanamo)
```

## Fetching OSM data

We are now retrieving OSM data for the study area of our choice: here we are interested in the accessibility (by bike, but we will see this later) from the train station of Grenoble.

Let's get the bounding box of the train station of Grenoble and then fetch the data for a radius of 5000 meters around it:

```{r cache = TRUE}
bb <- osmdata::getbb("Gare, 38000 Grenoble, France")
lon <- mean(bb[1, ])
lat <- mean(bb[2, ])
res <- get_city(c(lon, lat), r = 5000)
```

We can see that the resulting object is a list of sf layers: 

```{r}
print(names(res))
```

Depending on the radius defined in the `maposm::get_city` function the result list contains either individual buildings (if r <= 5000)
or urban patches (if r > 5000). As a result we need to filter the resulting list to remove any possible `NULL` entry:

```{r}
res <- res[!unlist(lapply(res, is.null))]
```

## Defining a reference point

We need to define a reference point from which the travel times will be computed. Since we used the train station of Grenoble as a center point to fetch the data, we can use the center of the circle that includes all the data:

```{r}
start <- st_centroid(res$circle)
```

## Sampling points

We then need a set of destinations to compute the travel times to. To do so, we sample roughly 200 points on the road network:

```{r}
# x <- st_sample(rbind(res$road,res$street), size = 200, type = "regular" )
x <- st_sample(res$road, size = 200, type = "regular")
points <- st_sf(id = seq_along(x), geom = x)
plot(st_geometry(points))
```


## Making a map of the result

```{r}
mf_map(res$circle, col = "#f2efe9", border = NA, add = FALSE)
mf_map(res$green, col = "#c8facc", border = "#c8facc", lwd = .5, add = TRUE)
mf_map(res$water, col = "#aad3df", border = "#aad3df", lwd = .5, add = TRUE)
mf_map(res$railway, col = "grey50", lty = 2, lwd = .2, add = TRUE)
mf_map(res$road, col = "white", border = "white", lwd = .5, add = TRUE)
mf_map(res$street, col = "white", border = "white", lwd = .5, add = TRUE)
mf_map(res$building, col = "#d9d0c9", border = "#c6bab1", lwd = .5, add = TRUE)
mf_map(res$circle, col = NA, border = "#c6bab1", lwd = 4, add = TRUE)
mf_map(start, pch = 23, col = "red", cex = 2, add = T)
mf_map(points, add = T, col = "blue")
```

## Computing travel times between the reference point and the sampled points

To do so, we can use the `osrmTable` function from the `osrm` package:

```{r}
mat <- osrmTable(
  dst = start,
  src = points,
  osrm.profile = "bike"
)

points$durations <- mat$durations
```

### Alternative to OSRM for computing travel times

Alternatively, if having a Valhalla instance under the hand and if the data preparation step included elevation data, you can also use it to compute travel times and benefit from taking the elevation into account:

```{r}
library(valhallr)

s <- as.data.frame(st_coordinates(st_transform(start, "EPSG:4326")))
names(s) <- c("lon", "lat")

d <- as.data.frame(st_coordinates(st_transform(points, "EPSG:4326")))
names(d) <- c("lon", "lat")

durs <- sources_to_targets(
  s,
  d,
  costing = "bicycle",
  hostname = Sys.getenv("VH_HOSTNAME"),
  port = Sys.getenv("VH_PORT")
)

points$durations <- durs$time / 60
```

Note that we use environment variables to store the hostname and port of the Valhalla instance because this is our
private instance. You need to replace `Sys.getenv('VH_HOSTNAME')` and `Sys.getenv('VH_PORT')` by the actual values you want
to use.

## Computing the positions of image points

We can now compute the positions of the image points using the `dc_move_from_reference_point` function.
It takes as input the reference point, the other points, and the column name of the travel times in the `other_points` object.

This function will move the points by computing the reference speed (the average distance / time) and then moving the points further or closer to the reference point depending on the travel time to reach them.

```{r}
pos_result <- dc_move_from_reference_point(
  reference_point = start,
  other_points = points,
  duration_col_name = "durations"
)
```

We can call the `summary` function to obtain a summary of the resulting object:

```{r}
summary(pos_result)
```

We can also plot the resulting object to visualize the positions of the image points and how they have been moved:

```{r}
plot(pos_result)
```

## Creating the concentric circles

Since the cartogram will replace the geographic distance by time distance we can create concentric circles to represent the time distance from the reference point.
Internally, this is possible because the reference speed has been stored in the resulting object of the `dc_move_from_reference_point` function.

The `steps` parameter is a list of the time distances we want to represent in the cartogram, in the same unit as the travel times (here, minutes).

```{r}
circles <- dc_concentric_circles(
  pos_result,
  steps = list(5, 10, 15, 20, 25, 30)
)
circles
```

## Creating the distance cartogram

We first need to compute the bounding box of all the layers we want to deform using the interpolation grid to ensure that the interpolation grid covers the whole area of interest.

```{r}
bbox <- dc_combine_bbox(res)
```

We can then create the interpolation grid:

The `precision` parameters controls the size of the grid cells (higher is more precise, for example 0.5 generally gives a coarse result, 2 a satisfactory result
and 4 a particularly fine result). A precision of 2 is usually a good default value and this is what we choose to use here.


```{r}
igrid <- dc_create(
  source_points = pos_result$source_points,
  image_points = pos_result$image_points,
  precision = 2,
  bbox = bbox
)
```

Various information can be obtained from the interpolation grid:

```{r}
summary(igrid)
```

We can also plot the interpolation grid to visualize the source grid, the interpolated grid, the distance between the image points and the interpolated points, and the strength of the deformation.

```{r}
plot(igrid)
```

## Deforming the background layers

Since we have multiple layers to deform, we can use the `dc_interpolate_parallel` function to deform all the layers at once:

```{r}
deformed <- dc_interpolate_parallel(igrid, res)
```

## Mapping the result


```{r}
mf_map(deformed$circle, col = "#f2efe9", border = NA, add = FALSE)
mf_map(deformed$green, col = "#c8facc", border = "#c8facc", lwd = .5, add = TRUE)
mf_map(deformed$water, col = "#aad3df", border = "#aad3df", lwd = .5, add = TRUE)
mf_map(deformed$railway, col = "grey50", lty = 2, lwd = .2, add = TRUE)
mf_map(deformed$road, col = "white", border = "white", lwd = .5, add = TRUE)
mf_map(deformed$street, col = "white", border = "white", lwd = .5, add = TRUE)
mf_map(deformed$building, col = "#d9d0c9", border = "#c6bab1", lwd = .5, add = TRUE)
mf_map(deformed$circle, col = NA, border = "#c6bab1", lwd = 4, add = TRUE)
mf_map(igrid$interpolated_grid, col = NA, border = "#940000", lwd = .1, add = TRUE)
mf_map(start, pch = 16, col = "red", cex = 2, add = TRUE)
mf_map(sf::st_intersection(circles, sf::st_union(sf::st_buffer(igrid$interpolated_grid, 0))), col = "red", lty = 2, lwd = 2, add = TRUE)

for (i in 1:4) {
  mapsf:::shadowtext(
    sf::st_bbox(circles$geometry[i])$xmax,
    start$geometry[[1]][2],
    paste(i * 5, "min"),
    col = "purple",
    bg = "white",
    r = 0.15
  )
}
```

If needed, we can add the various points (or a sample of them) and their corresponding travel times
to see that the deformation has been correctly applied:

```{r eval = FALSE}
mvd_pts <- igrid$interpolated_points
mvd_pts$durations <- c(0, points$durations)

mf_map(deformed$circle, col = "#f2efe9", border = NA, add = FALSE)
mf_map(deformed$green, col = "#c8facc", border = "#c8facc", lwd = .5, add = TRUE)
mf_map(deformed$water, col = "#aad3df", border = "#aad3df", lwd = .5, add = TRUE)
mf_map(deformed$railway, col = "grey50", lty = 2, lwd = .2, add = TRUE)
mf_map(deformed$road, col = "white", border = "white", lwd = .5, add = TRUE)
mf_map(deformed$street, col = "white", border = "white", lwd = .5, add = TRUE)
mf_map(deformed$building, col = "#d9d0c9", border = "#c6bab1", lwd = .5, add = TRUE)
mf_map(deformed$circle, col = NA, border = "#c6bab1", lwd = 4, add = TRUE)
mf_map(igrid$interpolated_grid, col = NA, border = "#940000", lwd = .1, add = TRUE)
mf_map(start, pch = 16, col = "red", cex = 2, add = TRUE)
mf_map(sf::st_intersection(circles, sf::st_union(sf::st_buffer(igrid$interpolated_grid, 0))), col = "red", lty = 2, lwd = 2, add = TRUE)

for (i in 1:4) {
  mapsf:::shadowtext(
    sf::st_bbox(circles$geometry[i])$xmax,
    start$geometry[[1]][2],
    paste(i * 5, "min"),
    col = "purple",
    bg = "white",
    r = 0.15
  )
}

mf_map(mvd_pts, add = TRUE, col = "blue")
mf_label(mvd_pts, var = "durations", col = "black", halo = TRUE, overlap = TRUE)
```

## Conclusion

Distance cartograms are both a way of analyzing space and a way of representing it. 
In the case of *unipolar* cartograms, as proposed here, reading is generally fairly straightforward, making it easy to highlight portions of the territory that are difficult (or at least more difficult than the average) to reach for example.




